Model Finding for Recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples Utilisables par Isabelle et Coq'' (CUIC). }

### 2015-05-08 tex/body.tex
more

Unfortunately, none of these fragments can accommodate the vast majority of
quantified formulas that correspond to recursive function definitions: The
essentially uninterpreted fragment does not allow the argument of a
recursive function to be used inside a complex term on the right-hand side,
whereas the finite model finding techniques of Reynolds et al.\ are not
applicable for functions over infinite domains such as the integers or
algebraic datatypes. A simple example where both approaches fail is
\begin{equation}\label{eq:pii}
\forall x : {\typeint}.\;\, \lite\bigl(
x \leq 0,\allowbreak\; \con{p}( x ) \teq 1,\allowbreak\; \con{p}( x ) \teq 2 * \con{p}( x - 1 ) \bigr)
\end{equation}
This state of affairs is highly unsatisfactory, given the frequency of
recursive definitions in practice and the impending addition of a dedicated
command, \texttt{define-fun-rec}, to the SMT-LIB standard \cite{smtlib25}.

In this paper, we present a method for translating formulas involving recursive function
definitions into formulas where finite model finding techniques can be applied.
The recursive functions must meet a semantic criterion to be admissible
(Section~\ref{sec:prelim}). This criterion is met by well-founded (terminating)
recursive function definitions. It is not met by inconsistent
definitions such as $\forall x : {\typeint}.\;\, \con{f}(x) \teq \con{f}(x) +
1$. %ct and it is possible to construct consistent examples that do not meet it.
%as well as other example of consistent definition.

We define a translation for a class of
formulas involving admissible recursive function definitions
(Section~\ref{sec:encoding}). The main insight is that a
recursive definition $\forall x : \tau.\;\, \const{f}(x) \teq t$
can be translated to
$\forall a :
\fargtype{\tau}.\;\, \const{f}(\farg{\con{f}}(a)) \teq t[\concret(a)/x]$, where
$\fargtype{\tau}$ is an uninterpreted \relax{abstract} type and $\farg{\con{f}} :
\fargtype{\tau} \to \tau$ converts the abstract type to the concrete
type. The translation preserves satisfiability and, for admissible definitions,
unsatisfiability, and makes finite model finding possible for problems in this
class.

Our empirical evaluation on benchmarks from the IsaPlanner proof planner~\cite{DBLP:conf/itp/JohanssonDB10}
and the Leon verifier~\cite{blanc2013overview} provides
evidence that this translation improves the effectiveness of the SMT solvers
\cvc and \ziii for finding counterexamples to verification conditions
(Section~\ref{sec:evaluation}). The approach is implemented as a preprocessor
in \cvc (Section~\ref{sec:front-end}). When invoked with a specific
command-line option, the solver assumes that functions introduced using the
\texttt{define-fun-rec} command are admissible and translates them accordingly.

@@

A \emph{$\Sigma$-interpretation $\I$} %is a mathematical structure that
maps each type $\tau \in \stypes{\Sigma}$ to a nonempty set $\tau^\I$,
the \emph{domain} of~$\tau$ in~$\I$,
and each function symbol $\con{f} : \tau_1 \times \cdots \times \tau_n \rightarrow \tau$ in
$\sfuns{\Sigma}$
to a total function $\con{f}^\I : \tau_1^\I \times \cdots \times \tau_n^\I \rightarrow \tau^\I$.
A \emph{theory} is a pair $T = (\Sigma, \Mo)$ where
$\Sigma$ is a signature and $\Mo$ is a class of $\Sigma$-interpretations,
the \emph{models} of $T$.
A $\Sigma$-formula $\varphi$ is
\emph{$T$-satisfiable}
if it is satisfied by some interpretation in $\Mo$;
otherwise, it is \emph{$T$-unsatisfiable}.
A formula $\varphi$ \emph{$T$-entails} $\psi$, written $\varphi \models_T \psi$,
if all interpretations in $\Mo$ that satisfy $\varphi$ also satisfy $\psi$.
Two formulas $\varphi$ and $\psi$ are \emph{$T$-equivalent} 
if each $T$-entails the other.
Given a signature $\Sigma$,
the \emph{theory of equality with uninterpreted functions} \euf
consists of the set of all $\Sigma$-interpretations.
Informally, we refer to the type and function symbols in this theory as \emph{uninterpreted}.
%\ct{This is both ambiguous and insufficient.
%Since this is SMT we should fix some background theory $T$,
%then extend its signature with additional symbols and call \emph{those} uninterpreted.
%}

@@

\begin{definition}\rm
A formula $\varphi$ is in \emph{definitional form with respect to}
$\{ \con{f}_1, \ldots, \con{f}_n \} \subseteq \sfuns{\Sigma}$ if it is of the
form
%
%\kern-\abovedisplayskip
%\kern+\abovedisplayshortskip
%
$(\forallf{\con{f}_1} \vec x_1.\; \varphi_1) \wedge \cdots \wedge
(\forallf{\con{f}_n} \vec x_n.\; \varphi_n) \wedge \psi$,
%
where $\con{f}_1, \ldots, \con{f}_n$ are distinct function symbols,
%ct added
$\forallf{\con{f}_i} \vec x_i.\; \varphi_i$ is a function definition
for $i = 1, \ldots, n$,
%
%\ct{I think we need more conditions here on how a symbol $\con{f}_i$ can occur in $\varphi_j$.}
%\jb{do we? I thought that the admissibility criterion below is enough.}
% should be enough, unless I'm missing something? -ajr
and $\psi$ contains no function definitions.
We call $\psi$ the \emph{conjecture} of $\varphi$.
%\ct{Don't we need additional restrictions on the $\varphi_i$'s here?
%(E.g., they are quantifier-free, or perhaps existential or containing only let binders.)}
% Now fixed above -ajr
\end{definition}

@@

First-order logic has no built-in notion of computation or termination. To ensure
that a function specification is well-founded, it is sufficient to require that
the function would terminate when seen as a functional program, under \emph{some}
evaluation order. For example, definition~(\ref{eq:pii}) from the
introduction can be shown well-founded under an evaluation strategy that
evaluates the condition of a $\lite$ before evaluating the relevant branch,
ignoring the other branch. Logically, such dependencies can be captured by
congruence rules. Krauss developed these ideas in the more general context of
higher-order logic \cite[Section 2]{krauss-2009-phd}.

\begin{theorem}\label{thm:adm}
If\/
$\DDD$ is a set of well-founded function definitions for\/
$\sfundefs{\Sigma} = \{\con{f}_1,\ldots,\con{f}_n\}$, then it is admissible.
\end{theorem}
\begin{longv}
\begin{proofsketch}
Let $\psi$ be a satisfiable formula closed under function expansion with
respect to $\DDD$. We show that $\psi \wedge \bigwedge \Delta$ is also
satisfiable. Let $\I$ be a model of $\psi$, and
let $\I_0$ be the restriction of $\I$ to the function symbols in $\sfuns{\Sigma} -
\sfundefs{\Sigma}$. Because well-founded definitions uniquely characterize
the interpretation of the functions they define, there exists a
$\Sigma$-interpretation $\J$ that extends $\I_0$ such that $\J \models \Delta$.
%
Since $\psi$ is closed under function expansion, it already constraints the
functions in $\sfundefs{\Sigma}$ recursively as far as is
necessary for interpreting $\psi$. Thus, any point $v$ for which
$\const{f}_i^\I(v)$ is needed for interpreting $\psi$ will have its
expected value according to its definition and hence coincide with $\J$.
And since $\psi^\I$ does not depend on the interpretation at the other
points, $\J$ is, like $\I$, a model of $\psi$.
Since $\J \models \Delta$ by assumption, we have $\J \models \psi \wedge
\bigwedge \Delta$, as desired.
%We have
%$\varphi \,\models_T\, \bigwedge\nolimits_{\,i=1}^{n} \{ \varphi_i[ \vec t / \vec x ] \mid \con{f}_i( \vec t ) \in \terms( \varphi ) \}$.
\qed
\end{proofsketch}
\end{longv}

>>>>

Unfortunately, none of these fragments can accommodate the vast majority of
quantified formulas that correspond to recursive function definitions: The
essentially uninterpreted fragment does not allow the argument of a
recursive function to be used inside a complex term on the right-hand side,
whereas the finite model finding techniques of Reynolds et al.\ are not
applicable for functions over infinite domains such as the integers or
algebraic datatypes. A simple example where both approaches fail is
$\forall x : {\typeint}.\;\, \lite\bigl(
x \leq 0,\allowbreak\; \con{p}( x ) \teq 1,\allowbreak\; \con{p}( x ) \teq 2 * \con{p}( x - 1 ) \bigr)$.
This state of affairs is unsatisfactory, given the frequency of
recursive definitions in practice and the impending addition of a dedicated
command, \texttt{define-fun-rec}, to the SMT-LIB standard \cite{smtlib25}.

In this paper, we present a method for translating formulas involving recursive function
definitions into formulas where finite model finding techniques can be applied.
The recursive functions must meet a semantic criterion to be admissible
(Section~\ref{sec:prelim}). This criterion is met by well-founded (terminating)
recursive function definitions. \begin{longv}It is not met by inconsistent
definitions such as $\forall x : {\typeint}.\;\, \con{f}(x) \teq \con{f}(x) +
1$.\end{longv} %ct and it is possible to construct consistent examples that do not meet it.
%as well as other example of consistent definition.

We define a translation for a class of
formulas involving admissible recursive function definitions
(Section~\ref{sec:encoding}). The main insight is that a
recursive definition $\forall x : \tau.\;\, \const{f}(x) \teq t$
can be translated to
$\forall a :
\fargtype{\tau}.\;\, \const{f}(\farg{\con{f}}(a)) \teq t[\concret(a)/x]$, where
$\fargtype{\tau}$ is an uninterpreted \relax{abstract} type and $\farg{\con{f}} \begin{longv}:
\fargtype{\tau} \to \tau\end{longv}$ converts the abstract type to the concrete
type. The translation preserves satisfiability and\begin{longv}, for admissible definitions,\end{longv}
unsatisfiability, and makes finite model finding possible for problems in this
class.

Our empirical evaluation on benchmarks from the IsaPlanner proof planner~\cite{DBLP:conf/itp/JohanssonDB10}
and the Leon verifier~\cite{blanc2013overview} provides
evidence that this translation improves the effectiveness of the SMT solvers
\cvc and \ziii for finding counterexamples to verification conditions
(Section~\ref{sec:evaluation}). The approach is implemented as a preprocessor
in \cvc (Section~\ref{sec:front-end}).\begin{longv} When invoked with a specific
command-line option, the solver assumes that functions introduced using the
\texttt{define-fun-rec} command are admissible and translates them accordingly.\end{longv}

@@

A \emph{$\Sigma$-interpretation $\I$} %is a mathematical structure that
maps each type $\tau \in \stypes{\Sigma}$ to a nonempty set $\tau^\I$,
the \emph{domain} of~$\tau$ in~$\I$,
and each function symbol $\con{f} : \tau_1 \times \cdots \times \tau_n \rightarrow \tau$ in
$\sfuns{\Sigma}$
to a total function $\con{f}^\I : \tau_1^\I \times \cdots \times \tau_n^\I \rightarrow \tau^\I$.
A \emph{theory} is a pair $T = (\Sigma, \Mo)$ where
$\Sigma$ is a signature and $\Mo$ is a class of $\Sigma$-interpretations,
the \emph{models} of $T$.
A $\Sigma$-formula $\varphi$ is
\emph{$T$-satisfiable}
if it is satisfied by some interpretation in $\Mo$\begin{longv};
otherwise, it is \emph{$T$-unsatisfiable}\end{longv}.
A formula $\varphi$ \emph{$T$-entails} $\psi$, written $\varphi \models_T \psi$,
if all interpretations in $\Mo$ that satisfy $\varphi$ also satisfy $\psi$.
Two formulas $\varphi$ and $\psi$ are \emph{$T$-equivalent} 
if each $T$-entails the other.
Given a signature $\Sigma$,
the \emph{theory of equality with uninterpreted functions} \euf
consists of the set of all $\Sigma$-interpretations.
Informally, we refer to the type and function symbols in this theory as \emph{uninterpreted}.
%\ct{This is both ambiguous and insufficient.
%Since this is SMT we should fix some background theory $T$,
%then extend its signature with additional symbols and call \emph{those} uninterpreted.
%}

@@

\begin{definition}\rm
A formula $\varphi$ is in \emph{definitional form with respect to}
$\{ \con{f}_1, \ldots, \con{f}_n \} \subseteq \sfuns{\Sigma}$ if it is of the
form
%
%\kern-\abovedisplayskip
%\kern+\abovedisplayshortskip
%
$(\forallf{\con{f}_1} \vec x_1.\; \varphi_1) \wedge \cdots \wedge
(\forallf{\con{f}_n} \vec x_n.\; \varphi_n) \wedge \psi$,
%
where $\con{f}_1, \ldots, \con{f}_n$ are distinct\begin{longv} function symbols\end{longv},
%ct added
$\forallf{\con{f}_i} \vec x_i.\; \varphi_i$ is a function definition
for $i = 1, \ldots, n$,
%
%\ct{I think we need more conditions here on how a symbol $\con{f}_i$ can occur in $\varphi_j$.}
%\jb{do we? I thought that the admissibility criterion below is enough.}
% should be enough, unless I'm missing something? -ajr
and $\psi$ contains no function definitions.
We call $\psi$ the \emph{conjecture} of $\varphi$.
%\ct{Don't we need additional restrictions on the $\varphi_i$'s here?
%(E.g., they are quantifier-free, or perhaps existential or containing only let binders.)}
% Now fixed above -ajr
\end{definition}

@@

First-order logic has no built-in notion of computation or termination. To ensure
that a function specification is well-founded, it is sufficient to require that
the function would terminate when seen as a functional program, under \emph{some}
evaluation order. For example, the definition
$\forall x : {\typeint}.\;\, \lite\bigl(
x \leq 0,\allowbreak\; \con{p}( x ) \teq 1,\allowbreak\; \con{p}( x ) \teq 2 * \con{p}( x - 1 ) \bigr)$
can be shown well-founded under an evaluation strategy that
evaluates the condition of a $\lite$ before evaluating the relevant branch,
ignoring the other branch. Logically, such dependencies can be captured by
congruence rules. Krauss developed these ideas in the more general context of
higher-order logic \cite[Section 2]{krauss-2009-phd}.

\begin{theorem}\label{thm:adm}
If $\DDD$ is a set of well-founded function definitions for\/
$\sfundefs{\Sigma} \begin{longv} = \{\con{f}_1,\ldots,\con{f}_n\}\end{longv}$, then it is admissible.
\end{theorem}
\begin{longv}
\begin{proofsketch}
Let $\psi$ be a satisfiable formula closed under function expansion with
respect to $\DDD$. We show that $\psi \wedge \bigwedge \Delta$ is also
satisfiable. Let $\I$ be a model of $\psi$, and
let $\I_0$ be the restriction of $\I$ to the function symbols in $\sfuns{\Sigma} -
\sfundefs{\Sigma}$. Because well-founded definitions uniquely characterize
the interpretation of the functions they define, there exists a
$\Sigma$-interpretation $\J$ that extends $\I_0$ such that $\J \models \Delta$.
%
Since $\psi$ is closed under function expansion, it already constraints the
functions in $\sfundefs{\Sigma}$ recursively as far as is
necessary for interpreting $\psi$. Thus, any point $v$ for which
$\const{f}_i^\I(v)$ is needed for interpreting $\psi$ will have its
expected value according to its definition and hence coincide with $\J$.
And since $\psi^\I$ does not depend on the interpretation at the other
points, $\J$ is, like $\I$, a model of $\psi$.
Since $\J \models \Delta$ by assumption, we have $\J \models \psi \wedge
\bigwedge \Delta$, as desired.
%We have
%$\varphi \,\models_T\, \bigwedge\nolimits_{\,i=1}^{n} \{ \varphi_i[ \vec t / \vec x ] \mid \con{f}_i( \vec t ) \in \terms( \varphi ) \}$.
\qed
\end{proofsketch}
\end{longv}
